---
title: "Cluster observations"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7,
  fig.retina=2
)
```


## Introduction

Alternative to the model that first partitions the tracks into segments and then clusters these segments into behavioral states, observations can be directly clustered into behavioral states using a different model. This (mixture) model also incorporates the non-parametric Bayesian approach that analyzes binned data streams and uses a truncated stick-breaking prior to determine the optimal number of states, but does so in a single stage.

This modeling approach provides results similar to that of state-space and hidden Markov models, but does not assume any Markov property or include a mechanistic process that drives state transitions (e.g., a correlated random walk). This model simply clusters observations together based on the discretized data streams. Therefore, this method provides fast and flexible inference on behavioral states from telemetry data, where individuals are pooled together to make population-level estimates.



## Empirical example

This example will demonstrate the steps taken to prepare and analyze an empirical telemetry dataset for a single individual fisher *Martes pennanti* (`leroy`) available in the `move` package (LaPoint et al., 2013) for the estimation of observation-level behavioral states.

```{r load data, message=FALSE}
library(bayesmove)
library(move)
library(dplyr)
library(ggplot2)
library(lubridate)

# Load data
data(leroy)

# Check data structure
leroy
plot(leroy); lines(leroy)
# Convert from `Move` object to data frame
leroy2<- data.frame(leroy) %>% 
  mutate(id = "leroy") %>% 
  rename(date = study.local.timestamp)

```


### Prepare the data

To start off, we are going to calculate step lengths and turning angles from the data, as well as time intervals to determine the primary time step. Additionally, step length and turning angle distributions will be explored to determine how to bin these data streams. Observations will be filtered to only retain those recorded at the primary time step.

```{r prep data}
# Calculate step lengths (step), turning angles (angle), and time step (dt) 
leroy2<- prep_data(dat = leroy2, coord.names = c("utm.easting","utm.northing"), id = "id")

# View distributions of data streams
hist(leroy2$step)
hist(leroy2$angle)

# Which dt is most common?
sort(table(leroy2$dt), decreasing = TRUE)[1:20]  #15 min (900 s) is primary dt


#Round times w/in 2 min of 15 min interval
leroy3<- round_track_time(dat = leroy2, id = "id", int = 900, tol = 120, time.zone = "UTC")
sort(table(leroy3$dt), decreasing = TRUE)[1:20]


# Create list from data frame
leroy.list<- df_to_list(dat = leroy3, ind = "id")

# Filter observations to keep only those recorded at 15 min (900 s)
leroy.filt<- filter_time(dat.list = leroy.list, int = 900) %>% 
  bind_rows
```


Now, let's bin step lengths and turning angles for analysis by the mixture model.

```{r bin data}
# Define bin number and limits for turning angles
angle.bin.lims<- seq(from=-pi, to=pi, by=pi/4)  #8 bins

# Define bin number and limits for step lengths
dist.bin.lims<- quantile(leroy.filt$step, c(0,0.25,0.50,0.75,0.90,1), na.rm=T)  #5 bins


leroy.filt2<- discrete_move_var(leroy.filt, lims = list(dist.bin.lims, angle.bin.lims),
           varIn = c("step","angle"), varOut = c("SL","TA"))


# Only retain id and discretized step length (SL) and turning angle (TA) columns
leroy.sub<- subset(leroy.filt2, select = c(id, SL, TA))
```


### Run the model

With the data streams in the correct format, we can now run the model to cluster observations into behavioral states.

```{r run model}
set.seed(1)

# Define model params
alpha=0.1  #prior
ngibbs=1000  #number of Gibbs sampler iterations
nburn=ngibbs/2  #number of burn-in iterations
nmaxclust=7  #number of maximum possible states (clusters) present

# Run model
dat.res<- cluster_obs(dat=leroy.sub, alpha=alpha, ngibbs=ngibbs, nmaxclust=nmaxclust,
                      nburn=nburn)

# Inspect traceplot of log-likelihood
plot(dat.res$loglikel, type = "l")
```


### Visualize the results

```{r}
# Determine the MAP estimate of the posterior
MAP.iter<- get_MAP_internal(dat = dat.res$loglikel, nburn = nburn)

# Determine the likely number of behavioral states (i.e., what are the fewest states that account for >= 90% of all observations?)

theta<- dat.res$theta[MAP.iter,]
names(theta)<- 1:length(theta)
theta<- sort(theta, decreasing = TRUE)
theta %>% cumsum()  #first 3 states likely (represent 97.6% of all assigned states)

# Store cluster order for plotting and behavioral state extraction
ord<- as.numeric(names(theta))
```


It appears that 3 behavioral states are likely based on the estimates from the `theta` matrix. To confirm that these states represent biologically interpretable movement patterns, we will also inspect the state-dependent distributions for all tested clusters/states.

```{r}
# Extract bin estimates for each possible state from the `phi` matrix of the model results
behav.res<- get_behav_hist(dat = dat.res, nburn = nburn, ngibbs = ngibbs, nmaxclust = nmaxclust,
                           var.names = c("Step Length","Turning Angle"), ord = ord,
                           MAP.iter = MAP.iter)

# Plot state-dependent distributions 
ggplot(behav.res, aes(x = bin, y = prop, fill = as.factor(behav))) +
  geom_bar(stat = 'identity') +
  labs(x = "\nBin", y = "Proportion\n") +
  theme_bw() +
  theme(axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x.bottom = element_text(size = 12),
        strip.text = element_text(size = 14),
        strip.text.x = element_text(face = "bold")) +
  scale_fill_manual(values = c(viridis::viridis(3), "grey35","grey35",
                               "grey35","grey35"), guide = FALSE) +
  scale_y_continuous(breaks = c(0.00, 0.50, 1.00)) +
  scale_x_continuous(breaks = 1:8) +
  facet_grid(behav ~ var, scales = "free_x")
```

Based on the resulting distributions, it appears that the first 3 states make biological sense and can potentially be classified as 'Encamped', 'Exploratory', and 'Transit' states. This is because state 1 (Encamped) exhibits low step lengths and high turning angles (near $\pi$ radians), state 2 (Exploratory) is characterized by a relatively uniform distribution of step lengths and turning angles, and state 3 (Transit) is characterized by large step lengths and low turning angles (near 0 radians). **It should be noted that these broad classes of behavioral states likely include a variety of different behaviors that could not be discerned from these telemetry data alone.**

Next, we will assign these 3 states to the associated observations. For observations that were assigned to one of the other four states that will not be retained, they will be assigned a value of NA. Similarly, observations that were omitted from the analysis (that were not recorded at a 15 min time step) will also be labeled NA.


### Mapping behavioral states estimates

To map the state estimates, we will first need to extract the state assignments associated with the MAP estimate and then re-classify these values to match the modified order from the `theta` and `phi` matrices.

```{r}
# Extract MAP estimate of behavioral states
z<- factor(dat.res$z[[MAP.iter]])
levels(z)<- ord

# Relabel factor levels
leroy.filt2$state<- ifelse(as.character(z) == 1, "Encamped",
                           ifelse(as.character(z) == 2, "Exploratory",
                                  ifelse(as.character(z) == 3, "Transit",
                                         NA)))

# Identify previously filtered observations and merge into final data frame
ind<- setdiff(1:nrow(leroy3), leroy.filt2$obs)
omit.df<- leroy3[ind,] %>% 
  mutate(obs = ind,
         time1 = NA,
         SL = NA,
         TA = NA,
         state = NA)


leroy.filt3<- rbind(leroy.filt2, omit.df)
leroy.filt3<- leroy.filt3[order(leroy.filt3$obs),]


# Map behavioral states for Leroy
ggplot() +
  geom_path(data = leroy.filt3, aes(x=x, y=y), color="gray60", size=0.25) +
  geom_point(data = leroy.filt3, aes(x, y, fill=state), size=1.5, pch=21, alpha=0.7) +
  geom_point(data = leroy.filt3 %>%
               group_by(id) %>%
               slice(which(row_number() == 1)) %>%
               ungroup(), aes(x, y), color = "green", pch = 21, size = 3, stroke = 1.25) +
  geom_point(data = leroy.filt3 %>%
               group_by(id) %>%
               slice(which(row_number() == n())) %>%
               ungroup(), aes(x, y), color = "red", pch = 24, size = 3, stroke = 1.25) +
  scale_fill_viridis_d("Behavior", na.value = "grey50") +
  labs(x = "Easting", y = "Northing") +
  theme_bw() +
  theme(axis.title = element_text(size = 16),
        strip.text = element_text(size = 14, face = "bold"),
        panel.grid = element_blank()) +
  guides(fill = guide_legend(label.theme = element_text(size = 12),
                             title.theme = element_text(size = 14)))
```

In this map, the green circle represents the first location and the red triangle represents the final location. As with other models that estimate behavioral states, these results can be used for a variety of subsequent analyses, such as to determine behavior-specific habitat suitability (via RSFs/SSFs) or to calculate activity budgets.



## References

LaPoint S, Gallery P, Wikelski M, Kays R. (2013). Animal behavior, cost-based corridor models, and real corridors. *Landscape Ecology*, 28(8), 1615–1630. [doi:10.1007/s10980-013-9910-0](http://dx.doi.org/10.1007/s10980-013-9910-0).

LaPoint S, Gallery P, Wikelski M, Kays R. (2013). Data from: Animal behavior, cost-based corridor models, and real corridors. *Movebank Data Repository*. [doi:10.5441/001/1.2tp2j43g](http://dx.doi.org/10.5441/001/1.2tp2j43g).
